Synthesizing Verified Components for Cyber Assured Systems Engineering \thanks{DISTRIBUTION STATEMENT A. Approved for public release.}

### 2022-09-19 paper/intro.tex
latest revision

Our work is embedded in a substantial collection of tools, so it is
important to clarify the research contribution. We see the following
as the contributions of this paper:

\begin{compactitem}
  \item The language and semantics of \emph{code contracts} provide a means
  to join system-level contract analysis results (arising from model
  checking) with component implementation correctness results (arising
  from property-preserving code generation).

>>>>

Our work is embedded in a substantial collection of tools, so it is
important to clarify the research impact. We see the following
as the distinguishing contributions of this paper:

\begin{compactitem}
  \item The language and semantics of \emph{code contracts} provide a
  means to join system-level contract analysis results (arising from
  model checking) with component implementation correctness results
  (arising from property-preserving code generation). This provides a
  strong formal path connecting (A) top-level system cyber-resilience
  requirements, (B) specifications for high-assurance components inserted
  to mitigate against specific cyber-resilience problems, and (C) synthesis
  to a programming language with a verified compiler.

### 2022-09-19 paper/related-work.tex
latest revision

%% DO NOT REMOVE --- BUT FEEL FREE TO REVISE
%% Specifically requested by the reviewers to be included in this paper!!!
Another model based approach to secure system architectures follows a pattern similar to \brfcs\ with attestation by inserting pre-verified components into the architecture to give certain guarantees \cite{6209212}.
In that work, the data connections are annotated with security properties as are the components.
A secondary analysis determines if the components meet the security requirements of the data connections, and if not, inserts new security enhancing components in the connection.
The insertion algorithm considers time cost when inserting components to preserve worst-case execution times mandated by the system.
\brfcs\ has the ability to insert pre-verified components based on cyber-requirements, but does not currently have the ability to annotate connections with security properties or automatically infer where secure components need to be inserted. 
\TODO{not clear where our work has benefits compared to this.}

Assume-guarantee reasoning for compositional verification in reactive systems is well-studied \cite{10.1007/978-3-642-28891-3_13, 10.1145/2658982.2527272, 10.1007/978-3-319-17524-9_7}.
The first steps toward synthesis of leaf-component contracts determined if leaf-component contracts are realizable \cite{10.1007/978-3-319-17524-9_13, 10.1007/978-3-319-29613-5_7}.
These led to algorithms for component synthesis from Lustre models based on realizability \cite{katis2017synthesis, 10.1007/978-3-319-89963-3_10}.
The algorithms discover a transition relation that preserves the contract guarantees under all allowed inputs and can be directly transformed into the language intended for the system's implementation.
Two such algorithms are implemented in the \agr\ verification engine \cite{jkind}.
These synthesis algorithms take in a declarative specification that states \emph{what is to be computed}, and then they have to infer \emph{how to compute it} for the implementation.
Code contracts are imperative, by design, so the designer specifies directly how the computation of the high-assurance component takes place thereby avoiding the question of realizability entirely.
That said, the test contracts in this paper can be declarative, stating what is to be computed, if desired, and \agr\ can then prove if the code contracts accomplishes that.

@@

A fully verified compiler has been developed that takes Lustre and compiles
it into assembly code, specified and verified in Coq \cite{10.1145/3140587.3062358}.
The key is in combining infinite sequences of data-flow models with incremental
manipulation of memories akin to an imperative model. CompCert is used on the
backend to create the final rendered executable \cite{compcert}.
This work is quite similar to
ours in providing a verified compilation chain from Lustre. However,
there are significant differences: notably, they tackle a more
comprehensive subset of Lustre than our work. We have been able to
work effectively with a relatively simple core language, \eg, without
\emph{clocking} features.  This subset aligns with our contract language, and
allows us to generate behaviorally equivalent code so we can show that the
generated code meets its logical specification.  This, in turn,
can be used as a leaf-level claim in a system-wide correctness argument
conducted by compositional reasoning in AGREE.
\TODO{not clear where the CompCert work falls short compared to ours.}

>>>>

%% DO NOT REMOVE --- BUT FEEL FREE TO REVISE
%% Specifically requested by the reviewers to be included in this paper!!!
Another model based approach to secure system architectures follows a pattern similar to \brfcs\ with attestation by inserting pre-verified components into the architecture to give certain guarantees \cite{6209212}.
In that work, the data connections are annotated with security properties as are the components.
A secondary analysis determines if the components meet the security requirements of the data connections, and if not, inserts new security enhancing components in the connection.
The insertion algorithm considers time cost when inserting components to preserve worst-case execution times mandated by the system.
\brfcs\ has the ability to insert pre-verified components based on cyber-requirements, but does not currently have the ability to annotate connections with security properties or automatically infer where secure components need to be inserted.

Assume-guarantee reasoning for compositional verification in reactive systems is well-studied \cite{10.1007/978-3-642-28891-3_13, 10.1145/2658982.2527272, 10.1007/978-3-319-17524-9_7}.
The first steps toward synthesis of leaf-component contracts determined if leaf-component contracts are realizable \cite{10.1007/978-3-319-17524-9_13, 10.1007/978-3-319-29613-5_7}.
These led to algorithms for component synthesis from Lustre models based on realizability \cite{katis2017synthesis, 10.1007/978-3-319-89963-3_10}.
The algorithms discover a transition relation that preserves the contract guarantees under all allowed inputs and can be directly transformed into the language intended for the system's implementation.
Two such algorithms are implemented in the \agr\ verification engine \cite{jkind}.
These synthesis algorithms take in a declarative specification that states \emph{what is to be computed}, and then they have to infer \emph{how to compute it} for the implementation.
Code contracts are imperative, by design, so the designer specifies directly how the computation of the high-assurance component takes place thereby avoiding the question of realizability entirely.
That said, the test contracts in this paper can be declarative, stating what is to be computed, if desired, and \agr\ can then prove if the code contracts accomplishe that.

@@

A fully verified compiler has been developed that takes Lustre and compiles
it into assembly code, specified and verified in Coq \cite{10.1145/3140587.3062358}.
The key is in combining infinite sequences of data-flow models with incremental
manipulation of memories akin to an imperative model. CompCert is used on the
backend to create the final rendered executable \cite{compcert}.
This work is quite similar to
ours in providing a verified compilation chain from Lustre. However,
there are significant differences: notably, they tackle a more
comprehensive subset of Lustre than our work. We have been able to
work effectively with a relatively simple core language, \eg, without
\emph{clocking} features.  This subset aligns with our contract language, and
allows us to generate behaviorally equivalent code so we can show that the
generated code meets its logical specification.  This, in turn,
can be used as a leaf-level claim in a system-wide correctness argument
conducted by compositional reasoning in AGREE.

### 2022-09-19 paper/synthesis.tex
latest revision

When full computational power is needed, we can use the expressive
power of AGREE code contracts to write arbitrarily complex filter or
monitor operations, as we have already seen. Our formal model of AGREE
syntax and semantics is applied to extract code contracts and map them
to equivalent logic functions. In particular, given a code contract with
functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state
variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics
given in Section \ref{sec:code-contracts} can be applied to create a
function
\[
 \konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env} = \sem{{\Eqs}\cdot G}
\]
Assume ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially
imperative. Then we can direcctly create a corresponding logic function
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
that takes input and state as parameters and returns an updated state
and output. Thus the environment $\mathit{Env}$ holding all previous
values of each variable can be replaced by \emph{parameters}, each of
which is able to hold only one data element. Assuming
$\mathit{Env}' = \konst{strmStep}\;t\; \mathit{Env}$, we can formally prove that
\[
\vdash (\mathit{Env}' \vert_{\mathbb{S}},
 \mathit{Env}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{Env}\vert_{\mathbb{I}(t)},
                   \mathit{Env}\vert_{\mathbb{S}(t)})
\]
In other words, projecting out the state and output components from
$\konst{strmStep}\;t\; \mathit{Env}$ equals the result of
applying \konst{stateStep} on the input and state components of
$\mathit{Env}$ at time $t$. This theorem connects the stream semantics
of the code contract to a pure logic function over the values
appearing in the stream. From this point we can use tools in the
CakeML eco-system to jump from the setting of higher order logic to
CakeML abstract syntax \cite{cakeml-translator}. Thus we have moved
from stream semantics to logic function to CakeML evaluation semantics
in a property-preserving way. Once in CakeML, further necessary
features can be added, namely imperative variables for storing state
between component invocations, also invocations of the (verified) foreign
function interface of CakeML \cite{cakeml-monadic}. Finally, the CakeML compiler can be
applied to the program. It delivers an executable which can be shown
to preserve the original contract.

@@

\item New values for the state variables are computed, in dependency
  order. At this point variables defined by \emph{followd-by} are
  dealt with. Suppose the variable definitions have the following
  form:
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
In the generated function, for the first invocation of \konst{stateStep} only,
the initializations are executed in order:
\[
  v_1 = i_1; \cdots ;  v_n = i_n;
\]
In all later invocations, the \emph{non-initialization} assignments are performed, also in order:
\[
  v_1 = e_1; \cdots ; v_n = e_n; \\
\]

>>>>

When full computational power is needed, we can use the expressive
power of AGREE code contracts to write arbitrarily complex filter or
monitor operations, as we have already seen. Our formal model of AGREE
syntax and semantics is applied to extract code contracts and map them
to equivalent logic functions. In particular, given a code contract
with functionality ${\Eqs}\cdot G$, input ports $\mathbb{I}$, state
variables $\mathbb{S}$, and output ports $\mathbb{O}$, the semantics
given in Section \ref{sec:code-contracts} can be applied to define a
stream-oriented ``step'' function
$\konst{strmStep} : \mathbb{N}_0 \to \mathit{Env} \to \mathit{Env}$ defined as
\[
 \konst{strmStep}\;t \; E = \sem{\mathit{Eqs} \cdot G}^E_t
\]
Now suppose ${\Eqs}\cdot G$ is \emph{Pascalish}, \ie, sequentially
imperative. Then we can direcctly create a logic function
\[
\konst{stateStep} : \mathit{input} \times \mathit{state} \to \mathit{state} \times \mathit{output}
\]
corresponding to \konst{strmStep} that takes input and state as
parameters and returns an updated state and output. Thus the
environment $\mathit{Env}$ holding all previous values of each
variable can be replaced by \emph{parameters}, each of which is able
to hold only one data element. Assuming $\mathit{E}'
= \konst{strmStep}\;t\; \mathit{E}$, we can formally prove that
\[
\vdash (\mathit{E}' \vert_{\mathbb{S}},
 \mathit{E}' \vert_{\mathbb{O}}) =
\konst{stateStep} (\mathit{E}\vert_{\mathbb{I}(t)},
                   \mathit{E}\vert_{\mathbb{S}(t)})
\]
In other words, projecting out the state and output components from
the result of $\konst{strmStep}\;t\; \mathit{E}$ equals the result of
applying \konst{stateStep} on the input and state components of
$\mathit{E}$ at time $t$. This theorem connects the stream semantics
of the code contract to a pure logic function over the values
appearing in the stream. From this point we can use tools in the
CakeML eco-system to jump from the setting of higher order logic to
CakeML abstract syntax \cite{cakeml-translator}. In this way we have
moved from stream semantics to logic function to CakeML evaluation
semantics in a property-preserving way. Once in CakeML, further
necessary features can be added, namely imperative variables for
storing state between component invocations, also invocations of the
(verified) foreign function interface of
CakeML \cite{cakeml-monadic}. Finally, the CakeML compiler can be
applied to the program. It delivers an executable which can be shown
to preserve the original contract.

@@

\item New values for the state variables are computed, in dependency
  order. At this point variables defined by \emph{followed-by} are
  dealt with. Suppose the variable definitions have the following
  form:
\[
\begin{array}{l}
  v_1 = i_1 \longrightarrow e_1 \\
  \cdots \\
  v_n = i_n \longrightarrow e_n \\
\end{array}
\]
In the generated function, for the first invocation of \konst{stateStep} only,
the initializations are executed in order:
\[
  v_1 = i_1; \cdots ;  v_n = i_n;
\]
In all later invocations, the \emph{non-initialization} assignments are performed, also in order:
\[
  v_1 = e_1; \cdots ; v_n = e_n; \\
\]
